CODE CHECK

### 2020-09-15 codecheck-paper.Rmd
On a roll...

# Background

@@

- eLife
- GigaScience
- AH, SciData
- Mozilla
- PLOS?

>>>>

# Open questions

* Can we switch to latex for this?  (Find it slightly easier.)
* Which journal to target -- maybe worry abotu that once we have a
  preprint.  F1000R my current favourite?

# Introduction

Many areas of scientific research now use computation to either
simulate or analyse their data.  These computations are increasingly
complex, and difficult to explain coherently in a paper. (MARWICK).
To complement the traditional route of sharing research by writing
papers, there is a growing trend/demand to share the underlying artifacts,
notably code and datasets, so that others can inspect, reproduce or
expand that work (Figure 1).  Some of the earliest proponents were
Buckheit and Donoho (1995) who commented "advert is scholarship for
the paper". (get correct quote).

So, assuming that researchers begin to share more artifacts, how
might they be examined to test that they do what they claim?  For
example, most scientific journals now require a data sharing statement
that outlines what data the authors have (or will) share.  The
implementation status of this varies considerably according to the
journal.  At one of the spectrum, there are specialist journals  that
have been created to accept "data papers" (e.g. *Scientific Data*,
... NAME COUPLE OF OTHERS).  These journals have established rigorous
procedures by which the data is validated according to standards in
each field.  At the other end of the spectrum, we still witness today
the infamous statement "Data available upon reasonable request"
whereby authors, well-intentioned, often cannot provide the data when
readers ask for it.

Given that there are no clear standards yet for sharing data, what
hope might there be for sharing computer programs?  Our anectodal
experiences on this matter suggest that often a variety of reasons are
given for why their code cannot be shared, e.g. "there is no
documentation / I do not want to maintain it / I do not want to give
away my code to my competitors".  Our view, rooted in open science
princples, is that wherever possible sharing code is good for the
community, as outlined ten years ago (Barnes, 2010).  Having the code
freely available, and suitably archived, provides a valuable resource
for others to learn from, even when it doesn't run or if there is no
documentation.  However, with a little effort, we believe that if
someone independent can run/use the code, this is worth documenting as
early as possible.  With this in mind, we have developed a set of
PRINCIPLES and an example workflow that provides what he hope is a
pragmatic way of checking that code works.  We call this system
CODECHECK as outlined next.

Figure 1: the inverse problem figure.

# What is a CODECHECK?

CODECHECK is best demonstrated by way of our example workflow, and
later we expand on the underlying principles.  The workflow involves
three groups of people (1) the AUTHOR providing the code to be
checked, (2) the PUBLISHER of a journal intereseted in publishing the
paper by the AUTHOR and (3) the CODECHECKER who checks that the
AUTHOR's code works.  The workflow that we have refined is documented
in Figure 2.

Step 1: the author submit their manuscript along with code and data
to the publishers.  The code and data need not be openly available at
this point.

Step 2: the publisher finds a codechecker to check the code.  This is
analgous to the publisher finding one or more reviewers to evaluating
the code.  However a cruical difference is that we suggest the
codechcker and the author can talk freely to each directly (see next).

Step 3: the codechecker runs the code, based on instructions provided
by the author, and checks if some/all of the results from the paper
can be reproduced.  (See later for exact nature of "reproduced").  If
there are any problems running the code, the codechecker asks the
author for help to resolve the problems.  The codechecker then tries
to run the code again.  This process iterates until either the
codechecker is successful, or the codechecker concludes the code does
not (*what do in this case*).  As part of this process, the
codechecker could work entirely locally on their own compute resource,
or in the cloud, e.g. using mybinder infrastructure.  Using such
cloud-based infrastructure allows for the codechecker and author to
collaboratively run the code together.

Step 4: The codechecker writes a certificate stating how the code was
run and includes a copy of outputs (figures, tables) that were
independently generated.

Step 5: the certificate and code/data get deposited on an open archive
(currently Zenodo), and the certificate is given to the PUBLISHER.

Step 6: the Publisher can then use the certificate as part of their
supplementary records for a paper, and give credit to the codechecker
for their work by depositing appropriate records (*bit hazy here*).

Figure 2: the detective.  Could we number the arrows as steps?
(Implementing the CODECHECK process)

## Variations

This example workflow leaves sufficient room for different
implementations of this process.  There are several aspects that can
be considered:

### When to do a codecheck

preprint (no publisher available), pre-condition for peer review,
during peer review, after acceptance.  Each of these have their own
pros/cons.

### Who does the codecheck

in house staff, pool of ECRs, volunteeres,

### Who knows who?

single blind, double blind, private/public communication between author and
codechecker, collaborative.  (This might need finessing with principle
2 below).

## Core principles

The workflow and variations just outlined below reflect our current
views on how a codecheck should be performed.  They are not set in
stone, but we do believe the following four core principles underlying
CODECHECK:

**1. Codecheckers record but don’t investigate or fix.**  The
codechecker follows the instructions provided by the author in running
the code.  If any steps are unclear, or if code does not run
correctly, the codechecker records this fact, and then communicates
this information to the author.  We believe that the job of the
codechecker is not to fix all these problems, but simply to report
them, and await a fix from the author.

**2. Communication between humans is key.**  Some code may just work
without any interaction (e.g. Manuel's certificate); however, more
often there are hidden dependencies that need to be found, or revised
software installations.  Allowing the codechecker and human to
communicate directly and openly with each other should make this
process as constructive as possible.  We feel that routing this
conversation (possibly anonymously) through a publisher would slow
things down and reduce the chances for community bulding.

**3. Credit is given to codecheckers.** (TODO Explain in a few
sentences what each means, e.g. see 1,2)

**4. Workflows must be auditable.** (TODO Explain in a few sentences
what each means, e.g. see 1,2)

# Register

We have a curated list of 15 certificates available at
<https://codecheck.org.uk/register>.  These are a mixture of
reproductions of papers for a variety of reasons (do we name them
all?) from interesting historical papers, to papers codechecked as
part of peer review, or because of public interest (covid).  several
covid models have been done during the time of peer review of these
papers, under our initative rather than requested from journal, but
certificate have then been acknowledged/cited in the paper (LANCET /
LSHTM).

We would like to briefly highlight three examples here.

1. Gigascience -- our first paper.  Only visualisation, not machine
   learning (which would have taken sseveral days of compute time).
2. report 9 -- covid model of UK lockdown model.
3. Geoscience

## Annotated certificate

Should we annotate a certificate, at least the cover page, to
demonstrate what we mean by a certificate?  We could print a four page
certificate 4-up so that it fits onto one page.  Text might be too
small though to read?  A 2-up, printed sideways would work well
though.

To run the system, we rely on github to host our work, and use a
custom R pacakge, codecheck, to help with the generation of
certificates and depositing work on zenodo.  (Cite zen4R).

All the related projectsfrom A-Z... e.g. CASCAID, Rescience, modeldb,
o2r,

# Limitations

Compute time:  for those papers that take significant compute time
(think days, not minutes), who will pay for the compute time?

Software: authors can currently provide code that requires
propieratiry software.  Given the prevalence of software like MATLAB,
this seems like a pragmatic choice, but then (a) we can't rely on
open infrastructure like mybinder, (b) codechecker must have that
software.

Can't someone cheat? yes.  we don't check it is correct code, just
simply that it runs.  Very low bar.  But by having the code open,
interested parties can then examine the code, and knowing that the
code will be open we hope will be an incentive for authors to share.

Who's got time for more peer review?  Agree, but ECRS might be
attracted to working with groups (recall it is not anonymous)

Who finds the codechecker?  Good point!  The journal.

As a principle, we do not require results to be exactly the same for a
codecheck to pass, simply that the code runs and generates the output
files that the author claims.  Stochastic simulations mean that often
we cannot get exactly the same, or even different versions of
libraries.  (Cite freesurfer example that generates different outputs
on different operating systems).

- untilhe co modelDB principle.

# Future work and conclusions

- possibility to allow CI for open papers such that for "nearly free"
  we can find out when softwave breaks, and if provenance tools are
  used, when.

- How to embed this in juornals.
- ???

@@

- eLife
- GigaScience
- AH, SciData
- Mozilla
- PLOS?
- Richard van Noorden, Dalmeet Singh (paper 9)
- Iain Davies and Sebastian 
- SSI for some funding.
